A Simple Solution of the Linearity Challenge

### 2025-12-28 paper/calculus.tex
tuning

The syntax of processes makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that each
message sent on one of the channels is forwarded to the other. As discussed in
the literature~\cite{LindleyMorris15}, this form is useful for modeling
higher-order sessions, that is the exchange of an existing channel on another
channel.
%
The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
of an empty message on the channel $x$. The latter process has no continuation,
while the former process continues as $P$ once the message has been received.
%
The process $\Fail\x$ can be used to denote a failure concerning the channel
$x$.
%
The process $\Case\x\z{Q_1}{Q_2}$ offers a choice on channel $x$ and continues
as either $Q_1$ or $Q_2$ with $z$ bound to the received continuation depending
on which branch is selected.
%
The process $\Select\x{\InTag_i}\z.P$ performs a choice (represented by a label
$\InTag_i$ with $i=1,2$) and sends a fresh continuation $z$ on the channel $x$.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of two fresh continuations channels $y$ and $z$ on the channel $x$. The receiver
can use $y$ and $z$ in whatever order. The sender forks into $P$ and $Q$, each
using $y$ and $z$ respectively. These forms are often used in conjunction with
links to model higher-order sessions.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

@@

The notions of free and bound channels are fairly standard, bearing in mind that
output prefixes bind continuation channels in (some) continuation processes. In
particular, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while
$\Fork\x\y\z{P}{Q}$ binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in
$P$.
%
We write $\fn{P}$ for the set of channels occurring free in $P$ and we identify
processes up to renaming of bound channels.

@@

Reduction is defined by the \rulename{r-*} rules.
%
The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists and \RConnect
are in one-to-one correspondence with the well-known reductions of session
calculi and the principal cut reductions of linear logic. These reductions erase
the topmost cut and replace it with zero, one or two new cuts, depending on the
number of continuations that are exchanged. Note that \RLink eliminates a link
$\Link\x\y$ by substituting $y$ for $x$ in the scope of $x$ and \RClose models
the communication of an empty message (without continuations), hence no new cut
is present in the reduct. \RSelect, \RFork and \RConnect are standard and
\RExists models the instantiation of a polymorphic variable $X$ as the
communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
replacing every free occurrence of the type variable $X$ with $A$.
%
The reduction rules \RWeaken and \RContract respectively represent the disposal
of an unused server and the duplication of a server. In these rules we use some
convenient (albeit slightly informal) notation for denoting sequences of
(pairwise distinct) channels and prefixes. In particular, $\seqof\z$ stands for
a sequence $z_1, \dots, z_n$ of channels, $\Weaken{\seqof\z}.Q$ stands for
$\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

>>>>

The syntax of processes makes use of an infinite set of \emph{channels}, ranged
over by $x$, $y$ and $z$, and is shown in~\Cref{tab:syntax}.
%
A link $\Link\x\y$ denotes the merging of the channels $x$ and $y$, so that each
message sent on one of the channels is forwarded to the other. As discussed in
the literature~\cite{LindleyMorris15}, this form is useful for modeling
higher-order sessions, that is the exchange of an existing channel on another
channel.
%
The processes $\Wait\x.P$ and $\Close\x$ respectively model the input and output
of an empty message on the channel $x$. The latter process has no continuation,
while the former process continues as $P$ once the message has been received.
%
The process $\Fail\x$ can be used to denote a failure concerning the channel
$x$.
%
The process $\Case\x\z{Q_1}{Q_2}$ offers a choice on channel $x$ and continues
as either $Q_1$ or $Q_2$ with $z$ bound to the received continuation channel
depending on which branch is selected.
%
The process $\Select\x{\InTag_i}\z.P$ performs a choice (represented by a label
$\InTag_i$ with $i=1,2$) and sends a fresh continuation $z$ on the channel $x$.
%
The processes $\Join\x\y\z.R$ and $\Fork\x\y\z{P}{Q}$ describe the input/output
of two fresh continuations channels $y$ and $z$ on the channel $x$. The receiver
can use $y$ and $z$ in whatever order. The sender forks into $P$ and $Q$, each
using $y$ and $z$ respectively. These forms are often used in conjunction with
links to model higher-order sessions.
%
The processes $\Forall\x\X\z.P$ and $\Exists\x{A}\z.P$ describe the input/output
of a type on the channel $x$ along with a fresh continuation $z$.

@@

The notions of free and bound channels are fairly standard, bearing in mind that
output prefixes bind continuation channels in (some) continuation processes. For
instance, $\Select\x{\InTag_i}\z.P$ binds $z$ in $P$, while $\Fork\x\y\z{P}{Q}$
binds $y$ in $P$ but not in $Q$ and $z$ in $Q$ but not in $P$.
%
We write $\fn{P}$ for the set of channels occurring free in $P$ and we identify
processes up to renaming of bound channels.

@@

Reduction is defined by the \rulename{r-*} rules.
%
The reduction rules \RLink, \RClose, \RSelect, \RFork, \RExists and \RConnect
are in one-to-one correspondence with the well-known reductions of session
calculi and the principal cut reductions of linear logic. These reductions erase
the topmost cut and replace it with zero, one or two new cuts, depending on the
number of continuation channels that are exchanged. Note that \RLink eliminates
a link $\Link\x\y$ by substituting $y$ for $x$ in the scope of $x$ and \RClose
models the communication of an empty message (without continuations), hence no
new cut is present in the reduct. \RSelect, \RFork and \RConnect are standard
and \RExists models the instantiation of a polymorphic variable $X$ as the
communication of a type $A$. We write $Q\subst{A}\X$ for the process obtained by
replacing every free occurrence of the type variable $X$ with $A$.
%
The reduction rules \RWeaken and \RContract respectively represent the disposal
of an unused server and the duplication of a server. In these rules we use some
convenient (albeit slightly informal) notation for denoting sequences of
(pairwise distinct) channels and prefixes. In particular, $\seqof\z$ stands for
a sequence $z_1, \dots, z_n$ of channels, $\Weaken{\seqof\z}.Q$ stands for
$\Weaken{z_1}\dots\Weaken{z_n}.Q$ and
$\Contract{\seqof\z}{\seqof\z'}{\seqof\z''}.R$ stands for
$\Contract{z_1}{z_1'}{z_1''}\dots\Contract{z_n}{z_n'}{z_n''}.R$.

### 2025-12-28 paper/conclusion.tex
tuning

The compact formalisation of \LCC is a good starting point for further
developments. We have already extended \LCC with support for infinite types and
recursive processes (the source code of this extension is publicly available in
the same repository as the one that contains the formalisation of
\LCC~\cite{PadovaniRaffaelli25}). In the future, it would be interesting to
formalise the strong normalisation property of \LCC as a consequence of cut
elimination of classical linear logic.

In this work we have focused on models and theories of \emph{binary sessions}
(those connecting exactly two processes), but there are also formalisations of
\emph{multiparty sessions}, notably those by \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which can be significantly more complex than
those of binary sessions. The formalisation by \citet{JacobsBalzerKrebbers22a}
amounts to 173kb and the one by \citet{TiroreBengstonCarbone25} to more than 1Mb
of Coq code. Also in these cases, the formalisation based on (virtual)
continuations~\citep{JacobsBalzerKrebbers22a} happens to be substantially
smaller. Whether this is a coincidence or further evidence of the effectiveness
of the continuation-based approaches is left as future work.

>>>>

The compact formalisation of \LCC is a good starting point for further
developments. We have already extended \LCC with support for coinductive (\ie
possibly infinite) types and recursive processes (the source code of this
extension is publicly available in the same repository of
\LCC~\cite{PadovaniRaffaelli25}). In the future, it would be interesting to
formalise the strong normalisation property of \LCC as a consequence of cut
elimination of classical linear logic.

In this work we have focused on models of \emph{binary sessions} (those
connecting exactly two processes), but there are also formalisations of
\emph{multiparty sessions}, notably those by \citet{JacobsBalzerKrebbers22a} and
\citet{TiroreBengstonCarbone25}, which can be significantly more complex than
those of binary sessions. The formalisation by \citet{JacobsBalzerKrebbers22a}
amounts to 173kb and the one by \citet{TiroreBengstonCarbone25} to more than 1Mb
of Coq code. Also in these cases, the formalisation based on (virtual)
continuations~\citep{JacobsBalzerKrebbers22a} happens to be substantially
smaller. Whether this is a coincidence or further evidence of the effectiveness
of the continuation-based approaches is left as future work.

### 2025-12-28 paper/introduction.tex
tuning

In this paper we approach the linearity challenge from a different angle:
\emph{instead of proposing new techniques that make it easy to formalise the
calculus in the challenge, we propose a (relatively) new calculus that is easy
to formalise with the existing techniques}.
%
More specifically, we describe the Linear Calculus of Continuations (\LCC) whose
type system coincides with the proof system of classical linear logic and that
features \emph{linear channels} instead of sessions. Unlike session endpoints,
which can be used multiple times albeit sequentially, linear channels are meant
to be used \emph{once}. \LCC retains the expressiveness of modeling structured
conversations using \emph{explicit continuations}, just like in the well-known
encoding of (binary) sessions into the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}:
each message exchanged on a linear channel may include one or two fresh channels
-- the \emph{continuations} -- on which the rest of the conversation takes
place.
%
Overall, \LCC is nothing but a low-level version of \CP~\cite{Wadler14} --
Wadler's calculus of sessions based on classical linear logic -- such that
sessions are encoded instead of being featured natively.

>>>>

In this paper we approach the linearity challenge from a different angle:
\emph{instead of proposing new techniques that make it easy to formalise the
calculus in the challenge, we propose a (relatively) new calculus that is easy
to formalise with the existing techniques}.
%
More specifically, we describe the Linear Calculus of Continuations (\LCC) whose
type system coincides with the proof system of classical linear logic and that
features \emph{linear channels} instead of sessions. While a session endpoint
can be used \emph{multiple times} (sequentially), linear channels must be used
\emph{exactly once}. \LCC retains the expressiveness of other session calculi
using \emph{explicit continuations}, as in the well-known encoding of (binary)
sessions into the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}:
each message exchanged on a linear channel may include one or two fresh channels
-- the \emph{continuations} -- on which the rest of the conversation takes
place.
%
Overall, \LCC is nothing but a low-level version of \CP~\cite{Wadler14} --
Wadler's calculus of sessions based on classical linear logic -- such that
sessions can be encoded instead of being featured natively.
